---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Microbiome Genus Set Enrichment Analysis (mgsea)

<!-- badges: start -->
<!-- badges: end -->

The goal of `mgsea` is to enable Genus Set Enrichment Analysis (GSEA)[^1] for microbiome data, allowing researchers to identify associations between microbial taxa and sample metadata features. This package provides a streamlined workflow for performing GSEA on microbiome datasets, leveraging the `phyloseq` object structure and the `fgsea` algorithm.

[^1]: Honestly, GSEA is a computational method that determines whether a predefined set of genes (or in this case, samples) shows statistically significant, concordant differences between two biological states. In the context of microbiome data, we adapt this concept to analyze whether certain sample groups are enriched in samples with high or low abundance of a target taxon.

## Installation

You can install the development version of `mgsea` from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("SyntheticCommunity/mgsea")
```

## Example Dataset

This is a basic example which shows you how to use this package:

```r
library(mgsea)
```

Initialize a test `phyloseq` object:

```{r}
#| include: false
if (Sys.getenv("HOME") == "/Users/gaoch") {
  devtools::load_all()
}
```


```{r}
library(phyloseq)
ps = ps_test_data
ps
```

Check `ps` is correctly loaded:

1. OTU table

    ```{r}
    otu_table(ps)
    ```

2. Taxonomy table

    ```{r}
    tax_table(ps)
    ```

3. Sample data

    ```{r}
    sample_data(ps)
    ```


## Overview

We can now visualize the abundance of all genera across samples:

```{r}
library(ggplot2)
ps_genus = tax_glom(ps, "Genus")
ps_genus = transform_sample_counts(ps_genus, function(x) x / sum(x))
plot_bar(ps_genus, fill = "Genus") +
    scale_y_continuous(labels = scales::percent_format())
```

```{r}
library(dplyr)

ps_data = psmelt(ps_genus)

sample_order_by_genus1 = ps_data |> 
    filter(Genus == "Genus1") |> 
    arrange(desc(Abundance)) |>
    pull(Sample)

ps_sorted = ps_data |> 
    mutate(Sample = factor(Sample, levels = sample_order_by_genus1))

ggplot(ps_sorted, aes(x = Sample, y = Abundance, fill = Genus)) +
    geom_bar(stat = "identity") +
    scale_y_continuous(labels = scales::percent_format()) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```


When we plot the abundance of "Genus1" across samples, we can see that it is more abundant in "Treat" group than "Ctrl" group:

```{r}
ps_genus1 = subset_taxa(ps_genus, Genus == "Genus1")
ps_data = psmelt(ps_genus1)

p_value = wilcox.test(Abundance ~ Group, data = ps_data)$p.value

ggplot(ps_data, aes(x = Group, y = Abundance)) +
    geom_boxplot() +
    geom_jitter(width = 0.2) +
    scale_y_continuous(labels = scales::percent_format()) +
    annotate("text", x = 1.5, y = max(ps_data$Abundance), label = paste0("p = ", signif(p_value, 2))) 
```


## Run Analysis

Run GSEA analysis:

```{r}
gsea_res <- run_microbiome_gsea(
    ps = ps,
    taxon_target = "Genus1",
    taxon_rank = "Genus",
    sample_feat = "Group"
)

gsea_res
```

## Visualization

Plot a single GSEA curve:

```{r}
gsea_plot(gsea_res, "Group_Treat")
```

Plot Curves for all significant sets:

```{r}
gsea_plot_all(gsea_res, filter = FALSE, ncol = 1)
```
